---
title: "UBER - ANA 515 - Assignment week 8"
author: "Bolaji Olakanmi"
date: "12/12/2021"
output: 
  word_document: default
  html_document: default
always_allow_html: true


---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1.	Discuss the business problem/goal 
This project performs some basic analysis and data visualizations of Uber data from April 2014 to September 2014 in New York City. The business goal of this anaysis is to gain insights into the travel habits of NYC  Uber riders. 

#2.	identify where the dataset was retrieved from 
Uber data sets was retrieved from a series of zipped csv files on a Google page, linked off of the Data Flair website. 

```{r}
library(ggplot2)
library(ggthemes)
library(lubridate)
library(dplyr)
library(tidyr)
library(DT)
library(scales)
library(readr)
```
#3. identify the code that imported and saved your dataset in R 

The code below sets the working directory and imports the data.

```{r}
getwd()
setwd("C:/Users/bolaj/Documents")
getwd()
```
Color Palette: 

```{r}
colors=c("#CC1011","#cc3333","#99cc99","#333333","#003399","#0683c9","#cc0033")
```
#Importing Data:

R Code below uses the read_csv() command to import the data from each month's csv file into an appropriately named variable. Afterwards, an rbind() command was used to create a single data_2014 dataframe by combining together all of the monthly dataframes. 


```{r}
library(tidyverse)
apr_data <- read.csv("C:/Users/bolaj/Documents/uber-raw-data-apr14.csv", header=TRUE )
may_data <- read.csv("C:/Users/bolaj/Documents/uber-raw-data-may14.csv")  
jun_data <- read.csv("C:/Users/bolaj/Documents/uber-raw-data-jun14.csv") 
jul_data <- read.csv("C:/Users/bolaj/Documents/uber-raw-data-jul14.csv") 
aug_data <- read.csv("C:/Users/bolaj/Documents/uber-raw-data-aug14.csv") 
sep_data <- read.csv("C:/Users/bolaj/Documents/uber-raw-data-sep14.csv")
data_2014 <- rbind(apr_data,may_data,jun_data,jul_data,aug_data,sep_data,fill=TRUE)
```


#4.	describe your data set 
#5.	discuss any data preparation
#6.	discuss the modeling 


```{r}
View(data_2014)
colnames(data_2014)
```

This data_2014 data frame has rows and columns. Initially, the variables are Date$Time, Latitude, Longitude, and Base. The Date$Time variable contains both pieces of information that were then combined into one variable. The Latitude and Longitude give spatial coordinates on the Earth regarding the Uber trips. Bases gives a label of one of five base locations for these rides. Given the nature of all four variables, they do not lend themselves to summary statistics as none are quantitative variables. As such, primary analysis will occur from the data visualizations given below using a series of commands from ggplot2. 


```{r}
data_2014$Date.Time <- as.POSIXct(data_2014$Date.Time, format = "%m/%d/%Y %H:%M:%S")
colnames(data_2014)
data_2014$Time <- format(as.POSIXct(data_2014$Date.Time, format = "%m/%d/%Y %H:%M:%S"), format="%H:%M:%S")
colnames(data_2014)
data_2014$Date.Time <- ymd_hms(data_2014$Date.Time)
data_2014$day <- factor(day(data_2014$Date.Time))
data_2014$month <- factor(month(data_2014$Date.Time, label = TRUE))
data_2014$year <- factor(year(data_2014$Date.Time))
data_2014$dayofweek <- factor(wday(data_2014$Date.Time, label = TRUE))
data_2014$hour <- factor(hour(hms(data_2014$Time)))
data_2014$minute <- factor(minute(hms(data_2014$Time)))
data_2014$second <- factor(second(hms(data_2014$Time)))
colnames(data_2014)
```
#7.	produce and discuss the output 
#8.	provide explanation with any visuals


In our first bar chart, the Uber trips are plotted by hours in a day (using a 24-hour clock) after creating an "hour_data" object in R. The summarize function from the dplyr library is used to create a total. A data table is then created and viewed from the "hour_data" object. 

```{r}
hour_data <- data_2014 %>%
  group_by(hour) %>%
  dplyr::summarize(Total = n()) 

datatable(hour_data)
View(hour_data)

ggplot(hour_data, aes(hour, Total)) + 
  geom_bar( stat = "identity", fill = "steelblue") +
  ggtitle("Trips Every Hour") +
  theme(legend.position = "none") +
  scale_y_continuous(labels = comma)
```

The bar chart we have above, shows hours on the horizontal and totals on the vertical axes.The code from Data Flair also demonstrated a red outline to the bars as a feature of ggplot. In terms of analysis, the 5-6 pm hour has the most number of trips across all bases from April to September 2014. From a broader perspective, the number of Uber trips is at a minimum in the early morning hours from 2-4 am. This begins to increase from 4 am to about 8 am in the morning, drops off slightly at 9 am and holds constant to about 12 noon. Afterwards, the number of trips builds from 12 noon to the maximum at 5 pm. There is a slight decrease from 6 pm to 7 pm, holds constant from 7-9 pm, and then drops throughout the rest of the night into the early morning hours of 2-4 am. A reasonable assumption would that the two local maximums occur roughly during the morning and evening commuting times with work and school times. 



The code we have below, creates a "month_hour" object from data_2014. 
```{r}
month_hour <- data_2014 %>%
  group_by(month, hour) %>%
  dplyr::summarize(Total = n())

datatable(month_hour)
View(month_hour)

ggplot(month_hour, aes(hour, Total, fill = month)) + 
  geom_bar( stat = "identity") +
  ggtitle("Trips by Hour and Month") +
  scale_y_continuous(labels = comma)
```

This creates a bar chart like before; however, the bars have monthly data superimposed onto them. While colorful, the bar chart is a little more difficult to analyze. What we can discern is that the trip behavior of New Yorkers is roughly the same over these six months across the 24-hour time frame. All months have similar local minimums and maximums in trip behavior, and all have similiar increases and decreases in number of trips at roughly the same times of the day. 


This creates a "day_group object" from data_2014. This bar chart catalogs every day of the month on the horizontal axis with total number of trips on the vertical axis. 
```{r}
day_group <- data_2014 %>%
  group_by(day) %>%
  dplyr::summarize(Total = n()) 
datatable(day_group)
View(day_group)

ggplot(day_group, aes(day, Total)) + 
  geom_bar( stat = "identity", fill = "steelblue") +
  ggtitle("Trips Every Day") +
  theme(legend.position = "none") +
  scale_y_continuous(labels = comma)
```

Overall, this bar chart does not provide a lot of new information. The number of trips per day is roughly the same for an entire month. The minimum bar occurs on day 31, which is not surprising as only three of the six surveyed months has 31 days. Disregarding that bar, least number of Uber trips is on the first day of the month with the most number on day 30. 


This creates a "day_month_group" object that further stratifies the prior plot. In this plot, the trips per days of the month are subdivided into the months. 
```{r}
day_month_group <- data_2014 %>%
  group_by(month, day) %>%
  dplyr::summarize(Total = n())

ggplot(day_month_group, aes(day, Total, fill = month)) + 
  geom_bar( stat = "identity") +
  ggtitle("Trips by Day and Month") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = colors)
```

While difficult to compare trips by day of the month across any of the months, what can be seen in the month of September is a repeating pattern where the number of Uber trips builds to a local maximum and then drops off immediately to a local minimum the next day. With additional time and data from 2014, additional analysis would be helpful to see if the local maximums correspond to a particular day of the week (Thursday or Friday to relax with friends) or if there is any correlation between the number of trips and holidays of this time period (Easter, Memorial Day, Fourth of July, Labor Day).




This creates a "month_weekday object" from data_2014. The bar graph puts a day of the week (by month) on the horizontal and totals on the vertical.
```{r}
month_weekday <- data_2014 %>%
  group_by(month, dayofweek) %>%
  dplyr::summarize(Total = n())

ggplot(month_weekday, aes(month, Total, fill = dayofweek)) + 
  geom_bar( stat = "identity", position = "dodge") +
  ggtitle("Trips by Day of the Week and Month") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = colors)
```

I left this graph in as an example of visual clutter. It is extremely difficult to pick out a day of the week and follow it across all of the months, given 7 colors. Overall, there is a trend of more Uber trips as the months go along. In addition, this graph does provide a little more context to the prior graph as we identified day of the week as an item of additional study. What is curious to me is that the maximum number of Uber trips by day of the week is different from month to month. For example, Wednesday has the most number of trips in April. In May, more trips were on Friday and closely followed up by Thursday. Thursday is the most popular day of the week for June and July, with Friday overtaking it for August. September has Tuesday, Thursday, Friday, and Saturday all equally popular. Additional analysis would be needed to parse out these reasons, such as Uber trips related to work obligations or vacationing. 



This creates a "month_group object" from data_2014. The bar chart has each of the months on the horizontal axis with trip totals on the vertical axis.
```{r}
month_group <- data_2014 %>%
  group_by(month) %>%
  dplyr::summarize(Total = n()) 
datatable(month_group)

ggplot(month_group, aes(month, Total, fill = month)) + 
  geom_bar( stat = "identity") +
  ggtitle("Trips by Month") +
  theme(legend.position = "none") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = colors)
```

From this bar chart, we can see that the total number of Uber trips in New York City has increased from April to September. By heights of the bars, the biggest increase in Uber rides occurred in the month of September. 



This section creates a visualization using the base label on the horizontal axis with trip totals on the vertical axis. 
```{r}
ggplot(data_2014, aes(Base)) + 
  geom_bar(fill = "darkblue") +
  scale_y_continuous(labels = comma) +
  ggtitle("Trips by Bases")
```

Overall, there are 5 bases. Base B02617 had the most number of Uber trips followed by B02598. Base B02512 had the least number of trips. 



This plot has bases on the horizontal, separated by month. 
```{r}
ggplot(data_2014, aes(Base, fill = month)) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma) +
  ggtitle("Trips by Bases and Month") +
  scale_fill_manual(values = colors)
```

Base B02617 had the most number of Uber trips in the months of July, August, and September as compared to all of the other months of all the other bases. Base B02598 (identified in the last bar chart with the second most trips) had its most number of trips in the month of May. Base B02512 had pretty consistent number of Uber trips across all six months. Base B02764 had very few trips from April to July, a slight jump in August, and then a very dramatic increase in September. If more data was supplied, it would be interesting to see how the number of trips for this base in successive months compared to these six months. 



This plot has the trips per base separated by day of the week.
```{r}
ggplot(data_2014, aes(Base, fill = dayofweek)) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma) +
  ggtitle("Trips by Bases and Day of Week") +
  scale_fill_manual(values = colors)
```

Again, this type of chart is attempting to provide more in-depth analysis; however, the number of colors is challenging for the reader. For four out of the five bases, Thursday is the most popular day of the week for an Uber, followed by Friday. 


This section creates a heatmap using ggplot. On this particular heatmap, day of the month is on the horizontal axis with hour of the day on the vertical axis. While I will create them below and they are colorful, I find them a little difficult to read. Using this legend, the lighter the color, the more Uber trips.  
```{r}
day_and_hour <- data_2014 %>%
  group_by(day, hour) %>%
  dplyr::summarize(Total = n())

datatable(day_and_hour)

ggplot(day_and_hour, aes(day, hour, fill = Total)) +
  geom_tile(color = "white") +
  ggtitle("Heat Map by Hour and Day")
```

Overall, it appears that the lightest colors are around hours 16-18, or 4-6 pm. This probably corresponds to many folks leaving work at that time.


This heatmap continues with day of the month on the horizontal but uses month of the year on the vertical axis. 
```{r}
ggplot(day_month_group, aes(day, month, fill = Total)) +
  geom_tile(color = "white") +
  ggtitle("Heat Map by Month and Day")
```

This visualization shows lighter colors in the month of September, meaning that there were more Uber trips in this month than in other months. Likewise, April and May has the darkest colors, corresponding to fewer trips. 




This heatmap plots day of the week on the horizontal axis with months on the vertical axis. This has the same type of shading; lighter blues indicate more Uber trips than darker blues. 
```{r}
ggplot(month_weekday, aes(dayofweek, month, fill = Total)) +
  geom_tile(color = "white") +
  ggtitle("Heat Map by Month and Day of Week")
```

The least number of trips was on Sundays and Mondays in the months of April and May. The most number of trips was taken in the month of September. In the month of July, the most popular days were Tuesday through Thursday. Friday was the most popular day for August. 

**Heatmap by Month and Bases**

This heatmap plotted bases on the horizontal axis with months on the vertical axis. 
```{r}
month_base <-  data_2014 %>%
  group_by(Base, month) %>%
  dplyr::summarize(Total = n()) 

day0fweek_bases <-  data_2014 %>%
  group_by(Base, dayofweek) %>%
  dplyr::summarize(Total = n()) 

ggplot(month_base, aes(Base, month, fill = Total)) +
  geom_tile(color = "white") +
  ggtitle("Heat Map by Month and Bases")
```


This final heat map plotted bases by days of the week. 
```{r}
ggplot(day0fweek_bases, aes(Base, dayofweek, fill = Total)) +
  geom_tile(color = "white") +
  ggtitle("Heat Map by Bases and Day of Week")
```

This heatmap shows a progression to lighter blues during the week different week bases.



```{r}
min_lat <- 40.5774
max_lat <- 40.9176
min_long <- -74.15
max_long <- -73.7004

ggplot(data_2014, aes(x=Lon, y=Lat)) +
  geom_point(size=1, color = "blue") +
  scale_x_continuous(limits=c(min_long, max_long)) +
  scale_y_continuous(limits=c(min_lat, max_lat)) +
  theme_map() +
  ggtitle("NYC MAP BASED ON UBER RIDES DURING 2014 (APR-SEP)")
```

**Conclusion**
The goal of this  project was meant to practice R code from the Data Flair website, especially with regard to data visualizations available to us from ggplot2. From our analysis, we can conclude the data visualization we've obtained from our analysis has helped in better understanding Uber pick up movements in the city of New York for the year 2014.
